PTFO 8-8695: Two Stars, Two Signals, No Planet

### 2020-06-16 paper/ms.tex
Update ms.tex

\explain{NOTE: added citation to Bodman et al 2017 in this paragraph.}
These light curves have been phenomenologically classified as
``persistent flux dips'' or ``transient flux dips'', based on whether
their depths and durations show variability over the 90-day K2
campaigns \citep{stauffer_orbiting_2017}.  In the terminology of
\citet{stauffer_orbiting_2017}, these objects are morphologically
distinct from ``scallop shell'' light curves, and are present in stars
at more advanced evolutionary disk stages than the ``dipper'' stars
\citep{ansdell_young_2016,bodman_dippers_2017,cody_manyfaceted_2018}.
The persistent and transient flux dip stars all show angular dips that
are cannot be explained as the effects of starspots.  These stars
typically have the following things in common:
\begin{enumerate}[topsep=0.5ex,itemsep=-0.5ex,partopsep=1ex,parsep=1.5ex]
  \item They are weak-lined T Tauri stars\added{ (under the
    definition of {\it e.g.}, \citealt{fang_star_2009})}.
  \item The spectral type is M2 to M5 ({\it e.g.},
    \citealt{rebull_usco_2018},~Figure~20).
  \item The age is typically $\lesssim$ 100$\,$Myr.\footnote{ At
    present, the oldest observed ``scallops'' are in the Pleaides
    \citep{rebull_rotation_2016}. One of these, EPIC 211013604, might
    meet the ``persistent dip'' classification.  If so, it is \deleted{also} the
    oldest known. } 
  \item The light curves show shallow, angular dips, \deleted{usually}
    superposed on large-amplitude smooth variability \added{at or near
    1:1 synchronicity}. The \replaced{latter}{smooth variability} is
    interpreted as stellar rotation.
  \item The rotation is rapid, with a period that is usually between
    0.5 and 2.0 days.
  \item There is rarely any infrared excess that is detectable in the
    WISE data (never any W4 detection; only a few W3 detections).
  \item They sometimes show multiple dips per cycle.
  \item The dip depths, durations, and phases can vary over just a few
    cycles ({\it e.g.}, EPIC 204143627).
  \item The dip depths can change after flares.
  \item They are rare at a population level, with an occurrence rate
  \replaced{of $\lesssim 1\%$}{between 0.1--1\%}
   of young M2-M5 stars \citep{rebull_usco_2018}.
\end{enumerate}
The 10.76$\,$hr signal in \ptfo\ meets all of these criteria.
\replaced{This is the first connection of \ptfo\ with this class of
objects, likely because the TESS data enabled us to put the dips in
the context of the asymmetric out-of-dip modulation.} {While this was
in fact mentioned by \citet{stauffer_orbiting_2017}, the
asymmetric out-of-dip modulation shown by the TESS data solidifies the
connection.}

@@

Second, approximately 40-50\% of the transient flux dip stars
discovered in $\rho$~Oph and Upper~Sco show two Lomb-Scargle periods,
and so \replaced{are apparently}{could be} binaries
\citep[][Table~1]{stauffer_orbiting_2017}.  This is higher than the
main-sequence companion fraction of ${\rm
CF}_{0.1\text{-}0.5\,M_\odot}^{\rm MS} = 33 \pm 5\%$\added{, and so it
could be an indication that the transient flux dip phenomenon is
linked to stellar binarity}
\citep{henry_solar_2006,duchene_stellar_2013,winters_solar_2019}.
\added{This fraction might be even more significant given the
selection bias against high mass ratio binaries, in which the second
star would be too faint to detect a second period.}

\added{However, the association between transient flux dips and
binarity is not iron-clad.  Firstly, multiple photometric periods do
not necessarily imply binarity. For instance, circumstellar material
could exist around single stars inside, at, or wide of co-rotation.
If it did, it could eclipse with a period different from the stellar
rotation.
Additionally, } low-mass pre-main-sequence stars \deleted{however}
have been shown to \added{have} companion fractions up to twice as
high in dispersed clusters such as Upper~Sco and Taurus
\citep{kraus_mapping_2008,kraus_mapping_2011}.  A high-resolution
imaging survey would be interesting, to determine whether the
transient flux dip stars truly have distinct population-level binarity
properties relative to other young low-mass stars.

@@

\explain{NOTE: ``High-latitude accretion hotspots'' was moved to this
paragraph.}
The explanations that we view as not yet ruled out include {\it A)}
transiting clumps of \added{dust or }gas at the Keplerian co-rotation
radius; {\it B)} occultations of \replaced{starspots}{hotspots on the
stellar surface} by an optically thick disk \citep{zhan_complex_2019};
and {\it C)} high-latitude accretion hotspots passing behind the star
\citep{yu_tests_2015}.  \replaced{The evolution of \ptfo\
over}{Observations of \ptfo\ from} the past decade may offer some
hints.  \deleted{Specifically, }\ptfo's transition between having
none, one, and multiple dips per cycle seems important
\added{(Figure~\ref{fig:o_minus_c})}.  \deleted{The dips would then
need to be caused by material that was somehow disrupted from the
planet, but somehow remained co-orbital for an extended duration. This
seems implausible.} \added{ The H$\alpha$ excess observed by
\citet{johnskrull_h_2016} also seems like an important piece of
evidence.  }

\explain{NOTE: This paragraph is newly added, to support our view of the
high-latitude accretion hotspot being plausible for \ptfo.}
The high-latitude accretion hotspot scenario seems qualitatively
capable of explaining most of the characteristics of \ptfo, including
the fact that its H$\alpha$ equivalent width of 11.4$\,$\AA\
\citep{van_eyken_ptf_2012} places it near the ``classical/weak'' T
Tauri division \citep{briceno_cidaII_2019}, and the fact that its
excess H$\alpha$ component varies in velocity between $-200$ and
$+200$ km$\,$s$^{-1}$ \citep{johnskrull_h_2016}.  The lack of a
NIR-excess could be problematic, but does not eliminate the
possibility that a tenuous amount of dust or gas remains near the
star, and is able to accrete onto it.  We view the main problem with
this explanation to be the likelihood that it explains the population
of transient flux dips.  RIK-210 in particular has dips too deep
($\sim$20\%) and too asymmetric to be explained by occultations of a
hot spot \citep{david_transient_2017}.

\explain{NOTE: This paragraph has been expanded to clarify the
possibility of a hotspot being occulted by a disk.}
Explanations {\it A} and {\it B} (circumstellar material at the
co-rotation radius, and occultations of spots by a disk) have the
appeal that they are flexible enough to explain not only the transient
and persistent-dip stars, but also the ``scallop shells''
\citep{stauffer_orbiting_2017}.  \replaced{Despite this appeal, the
possibility of distinct mechanisms explaining these distinct
variability classes remains open.}{ In the spot occultation scenario
\citep{zhan_complex_2019}, the ``blobs'' that provide the necessary
brightness contrast are on the surface of the star, rather than at the
co-rotation radius. The observing geometry would require an inner ring
to be projected into a narrow band in front of the stellar disk, in
order to produce variations with sufficiently short timescales.  The
outer disk would also need to either not be present, or to be
misaligned enough and insufficiently flared, that it did not fall
along the line of sight.  This geometric requirement seems too
contrived to occur in even $\lesssim$1\% of young M dwarf systems. }

\added{Out of the available explanations, we therefore consider the
clumpy torus of optically thick dust or gas near the co-rotation
radius to be most plausible.  The partially ionized circumstellar
hydrogen would readily explain the excess H$\alpha$ emission, and
changes in the clumping geometry would lead to changes in the observed
eclipse shapes.  The long-term stability, punctuated by changes in dip
phases, might be understood as an accretion disk subject to strong
magnetic fields going through ``stalled'' states in which material is
trapped near co-rotation \citep{dangelo_accretion_2012}.  Physical
models of this scenario may need to consider such states, in addition
to dissociation of the dust population
\citep{millan-gabet_circumstellar_2007,dullemond_inner_2010}.
Finally, the question of whether the opacity in fact comes from dust
or partially-ionized gas ({\it e.g.}, via gas continuum, atomic lines,
or perhaps TiO molecular opacity) remains an additional issue for
future exploration.  }

>>>>

\explain{NOTE: added citation to Bodman et al 2017 in this paragraph.}
These light curves have been phenomenologically classified as
``persistent flux dips'' or ``transient flux dips'', based on whether
their depths and durations show variability over the 90-day K2
campaigns \citep{stauffer_orbiting_2017}.  In the terminology of
\citet{stauffer_orbiting_2017}, these objects are morphologically
distinct from ``scallop shell'' light curves, and are present in stars
at more advanced evolutionary disk stages than the ``dipper'' stars
\citep{ansdell_young_2016,bodman_dippers_2017,cody_manyfaceted_2018}.
The persistent and transient flux dip stars all show angular dips that
are cannot be explained as the effects of starspots.  These stars
typically have the following things in common:
\begin{enumerate}[topsep=0.5ex,itemsep=-0.5ex,partopsep=1ex,parsep=1.5ex]
  \item They are weak-lined T Tauri stars\added{ (under the
    definition of {\it e.g.}, \citealt{fang_star_2009})}.
  \item The spectral type is M2 to M5 ({\it e.g.},
    \citealt{rebull_usco_2018},~Figure~20).
  \item The age is typically $\lesssim$ 100$\,$Myr.\footnote{ At
    present, the oldest observed ``scallops'' are in the Pleaides
    \citep{rebull_rotation_2016}. One of these, EPIC 211013604, might
    meet the ``persistent dip'' classification.  If so, it is \deleted{also} the
    oldest known. } 
  \item The light curves show shallow, angular dips, \deleted{usually}
    superposed on large-amplitude smooth variability \added{at or near
    1:1 synchronicity}. The \replaced{latter}{smooth variability} is
    interpreted as stellar rotation.
  \item The rotation is rapid, with a period that is usually between
    0.5 and 2.0 days.
  \item There is rarely any infrared excess that is detectable in the
    WISE data (never any W4 detection; only a few W3 detections).
  \item They sometimes show multiple dips per cycle.
  \item The dip depths, durations, and phases can vary over just a few
    cycles ({\it e.g.}, EPIC 204143627).
  \item The dip depths can change after flares.
  \item They are rare at a population level, with an occurrence rate
  \replaced{of $\lesssim 1\%$}{between 0.1--1\%}
   of young M2-M5 stars \citep{rebull_usco_2018}.
\end{enumerate}
The 10.76$\,$hr signal in \ptfo\ meets all of these criteria.
\replaced{This is the first connection of \ptfo\ with this class of
objects, likely because the TESS data enabled us to put the dips in
the context of the asymmetric out-of-dip modulation.} {While this was
mentioned by \citet{stauffer_orbiting_2017}, the
asymmetric out-of-dip modulation shown by the TESS data solidifies the
connection.}

@@

Second, approximately 40-50\% of the transient flux dip stars
discovered in $\rho$~Oph and Upper~Sco show two Lomb-Scargle periods,
\replaced{and so are apparently}{a possible sign of binarity} \removed{binaries}
\citep[][Table~1]{stauffer_orbiting_2017}.  This is higher than the
main-sequence companion fraction of ${\rm
CF}_{0.1\text{-}0.5\,M_\odot}^{\rm MS} = 33 \pm 5\%$\added{, which
could be an indication that the transient flux dip phenomenon is
linked to stellar binarity}
\citep{henry_solar_2006,duchene_stellar_2013,winters_solar_2019}.
\added{This fraction might be even more significant given the
selection bias against high mass ratio binaries, in which the second
star would be too faint to detect a second period.}

\added{However, the association between transient flux dips and
binarity is not iron-clad. Firstly, multiple photometric periods do
not necessarily imply binarity. For instance, circumstellar material
could exist around single stars at multiple orbital distances, and not
just at the co-rotation radius.
This could cause eclipses with periods different from the stellar
rotation period.
Additionally, } low-mass pre-main-sequence stars \deleted{however}
have been shown to \added{have} companion fractions up to twice as
high in dispersed clusters such as Upper~Sco and Taurus
\citep{kraus_mapping_2008,kraus_mapping_2011}.  A high-resolution
imaging survey would be interesting, to determine whether the
transient flux dip stars truly have distinct population-level binarity
properties relative to other young low-mass stars.

@@

\explain{NOTE: ``High-latitude accretion hotspots'' was moved to this
paragraph.}
The explanations that we view as not yet ruled out include {\it A)}
transiting clumps of \added{dust or }gas at the Keplerian co-rotation
radius; {\it B)} occultations of \replaced{starspots}{hotspots on the
stellar surface} by an optically thick disk \citep{zhan_complex_2019};
and {\it C)} high-latitude accretion hotspots passing behind the star
\citep{yu_tests_2015}.  \replaced{The evolution of \ptfo\
over}{Observations of \ptfo\ over} the past decade may offer some
hints.  \deleted{Specifically, }\ptfo's transition between having
no dips, one dip, and multiple dips per cycle seems important
\added{(Figure~\ref{fig:o_minus_c})}.  \deleted{The dips would then
need to be caused by material that was somehow disrupted from the
planet, but somehow remained co-orbital for an extended duration. This
seems implausible.} \added{ The H$\alpha$ excess observed by
\citet{johnskrull_h_2016} also seems like an important piece of
evidence.  }

\explain{NOTE: This paragraph is newly added, to support our view of the
high-latitude accretion hotspot being plausible for \ptfo.}
The high-latitude accretion hotspot scenario seems 
capable of explaining most of the qualitative characteristics of \ptfo.
Some degree of accretion onto the star seems likely, given 
the H$\alpha$ equivalent width of 11.4$\,$\AA\
\citep{van_eyken_ptf_2012}, which is near the boundary separating ``classical'' and ''weak-lined'' T
Tauri stars \citep{briceno_cidaII_2019}, and the high-velocity
components of the observed H$\alpha$ line profile ($\pm\,200$\,km$\,$s$^{-1}$ \citep{johnskrull_h_2016}.
The non-detection of any NIR excess could be problematic, but does not eliminate the
possibility that a tenuous amount of dust and gas remains near the
star, and is able to accrete onto it.  We view the main problem with
this explanation seems to require rare circumstances,
making it unlikely to be correct for the entire population
of transient flux dips.  RIK-210 in particular has dips that seem too deep
($\sim$20\%) and too asymmetric to be explained by occultations of a
hot spot \citep{david_transient_2017}.

\explain{NOTE: This paragraph has been expanded to clarify the
possibility of a hotspot being occulted by a disk.}
Explanations {\it A} and {\it B} (circumstellar material at the
co-rotation radius, and occultations of spots by a disk) have the
appeal that they are flexible enough to explain not only the transient
and persistent-dip stars, but also the ``scallop shells''
\citep{stauffer_orbiting_2017}.  \replaced{Despite this appeal, the
possibility of distinct mechanisms explaining these distinct
variability classes remains open.}{ In the spot occultation scenario
\citep{zhan_complex_2019}, the ``blobs'' that provide the necessary
brightness contrast are on the surface of the star, rather than at the
co-rotation radius. The observing geometry would require an inner ring
to be projected into a narrow band in front of the stellar disk, in
order to produce variations with sufficiently short timescales.  The
outer disk would also need to either not be present, or to be
misaligned enough and insufficiently flared, that it did not obstruct
the line of sight to the star.  This geometric requirement seems too
contrived to occur in even $\lesssim$1\% of young M dwarf systems. }

\added{Thus, while several of the proferred explanations deserve further
scrutiny, we consider the
clumpy torus of optically thick dust or gas near the co-rotation
radius to be most plausible.  The partially ionized circumstellar
hydrogen would readily explain the excess H$\alpha$ emission, and
changes in the clumping geometry would lead to changes in the observed
eclipse shapes.  The long-term stability, punctuated by changes in dip
phases, might be understood as an accretion disk subject to strong
magnetic fields going through ``stalled'' states in which material is
trapped near co-rotation \citep{dangelo_accretion_2012}.  Physical
models of this scenario may need to consider such states, in addition
to dissociation of the dust population
\citep{millan-gabet_circumstellar_2007,dullemond_inner_2010}.
Finally, the question of whether the opacity in fact comes from dust
or partially-ionized gas ({\it e.g.}, via gas continuum, atomic lines,
or perhaps TiO molecular opacity) remains an additional issue for
future exploration.  }
